---
title: "Final project: Map of Happiness"
author: "Laura Louise Rokkjær and Sofie K. Lloyd Thomsen"
date: "01/05/2022 updated `r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
library(htmlwidgets)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
options(scipen=999)
```

# Get libraries and load data

## Activate libraries
```{r libraries, eval=FALSE, echo=TRUE}
library(rgdal)
library(raster)
library(leaflet) 
library(dplyr)
library(readr) 
library(tidyverse) 
```


## Load data from 2015-2019
First, we need to load our five .csv files from the World Happiness Report. 
The data can also be found on Kaggle following [this](https://www.kaggle.com/datasets/unsdsn/world-happiness) link. 

```{r load data, eval=FALSE, echo=TRUE}
# read the five data frames from 2015-2019
data_2015 <- read_csv("../input/2015.csv")
data_2016 <- read_csv("../input/2016.csv")
data_2017 <- read_csv("../input/2017.csv")
data_2018 <- read_csv("../input/2018.csv")
data_2019 <- read_csv("../input/2019.csv")
```


We then want to add a year column to the five data frames, so we know what year the data comes from. 

```{r add year column, eval=FALSE, echo=TRUE}
# 2015
data_2015 <- data_2015 %>% 
  mutate(Year = "2015")

# 2016
data_2016 <- data_2016 %>% 
  mutate(Year = "2016")

# 2017
data_2017 <- data_2017 %>% 
  mutate(Year = "2017")

# 2018
data_2018 <- data_2018 %>% 
  mutate(Year = "2018")

# 2019
data_2019 <- data_2019 %>% 
  mutate(Year = "2019")
```


## Get values
We then want to combine all the country names into one data frame.

```{r country names, eval=FALSE, echo=TRUE}
# get country names from 2015-2019
Countries <- c(data_2015$Country, 
               data_2016$Country, 
               data_2017$Country, 
               data_2018$`Country or region`, 
               data_2019$`Country or region`)
```


...And we do the same with the year columns. 

```{r years, eval=FALSE, echo=TRUE}
# get years from 2015-2019
Years <- c(data_2015$Year,
           data_2016$Year,
           data_2017$Year,
           data_2018$Year,
           data_2019$Year)
```


...And the same with HPI scores

```{r Happiness, eval=FALSE, echo=TRUE}
# get happiness scores from 2015-2019
HPI_scores <-  c(data_2015$`Happiness Score`,
             data_2016$`Happiness Score`,
             data_2017$Happiness.Score,
             data_2018$Score,
             data_2019$Score)
```


...And for GDP per capita.

```{r GDP per capita, eval=FALSE, echo=TRUE}
# get GDP from 2015-2019
GDP <- c(data_2015$`Economy (GDP per Capita)`, 
         data_2016$`Economy (GDP per Capita)`, 
         data_2017$`Economy..GDP.per.Capita.`, 
         data_2018$`GDP per capita`,
         data_2019$`GDP per capita`)
```


...And for family/social support.

```{r family/social support, eval=FALSE, echo=TRUE}
# get family values from 2015-2019
Family <- c(data_2015$Family,
            data_2016$Family,
            data_2017$Family,
            data_2018$`Social support`,
            data_2019$`Social support`)
```


...And for health.

```{r health scores, eval=FALSE, echo=TRUE}
# get health values from 2015-2019
Health <- c(data_2015$`Health (Life Expectancy)`,
            data_2016$`Health (Life Expectancy)`,
            data_2017$Health..Life.Expectancy.,
            data_2018$`Healthy life expectancy`,
            data_2019$`Healthy life expectancy`)
```


# Create new data frames
In order to get the data we need, we create four new data frames. All four data frames has the columns 'Countries', '2015', '2016', '2017', '2018', '2019' and 'mean_scores'. The values in the year columns and mean_scores varies from data frame to data frame, and reflect happiness (HPI), economy (GDP per capita), Social support (family) and health (life expectancy).


## HPI data frame
Now we create a data frame with the columns 'HPI_scores', 'Countries' and the five 'Years'. 

```{r HPI, eval=FALSE, echo=TRUE}
# bind columns 
HPI_df <- cbind(Countries, HPI_scores, Years)


# convert to data frame
HPI_df <- as.data.frame(HPI_df)

# convert 'Scores' to numeric
HPI_df$HPI_scores <- as.numeric(HPI_df$HPI_scores)

# pivot data frame from long to wide
HPI_wide <- HPI_df %>%
  # column names = years
  pivot_wider(names_from = Years,
              # rows = HPI scores
              values_from = HPI_scores,
              # if there is no HPI_scores write 'NA'
              values_fill = NA)

# get mean HPI score for all years
HPI_wide$mean_score = rowMeans(HPI_wide[,c(2:6)], na.rm = TRUE)

# round decimals to 2 digits
HPI_wide <- HPI_wide %>%  
  mutate_if(is.numeric,
            round,
            digits = 2)

print(HPI_wide)
```


## GDP data frame
Now we create a data frame with the columns 'GDP', 'Countries' and the five 'Years'. 

```{r GDP, eval=FALSE, echo=TRUE}
# bind columns
GDP_df <- cbind(Countries, GDP, Years)

# convert to data frame
GDP_df <- as.data.frame(GDP_df)

# convert GDP to numeric
GDP_df$GDP <- as.numeric(GDP_df$GDP)


# pivot data frame from long to wide
GDP_wide <- GDP_df %>%
  # column names = years
  pivot_wider(names_from = Years,
              # rows = GDP
              values_from = GDP,
              # if there is no GDP score, write 'NA'
              values_fill = NA)

# get mean GDP from all years
GDP_wide$mean_score = rowMeans(GDP_wide[,c(2:6)], na.rm = TRUE)

# round decimals to 2 digits
GDP_wide <- GDP_wide %>%  
  mutate_if(is.numeric,
            round,
            digits = 2)

print(GDP_wide)
```


## Family data frame
Now we create a data frame with the columns 'Family', 'Countries' and the five 'Years'.

```{r family, eval=FALSE, echo=TRUE}
# bind columns
Family_df <- cbind(Countries, Family, Years)

# convert to data frame
Family_df <- as.data.frame(Family_df)

# convert Family to numeric
Family_df$Family <- as.numeric(Family_df$Family)


# pivot data frame from long to wide
family_wide <- Family_df %>%
  # column names = years
  pivot_wider(names_from = Years,
              # rows = Family
              values_from = Family,
              # if there is no Family score, write 'NA'
              values_fill = NA)

# get mean family score of all years
family_wide$mean_score = rowMeans(family_wide[,c(2:6)], na.rm = TRUE)

# round decimals to 2 digits
family_wide <- family_wide %>%  
  mutate_if(is.numeric,
            round,
            digits = 2)

print(family_wide)
```


## Health data frame
Now we create a data frame with the columns 'Health', 'Countries' and the five 'Years'.

```{r health, eval=FALSE, echo=TRUE}
# cbind
health_df <- cbind(Countries, Health, Years)

# convert to data frame
health_df <- as.data.frame(health_df)

# convert Family to numeric
health_df$Health <- as.numeric(health_df$Health)

# pivot data frame from long to wide
health_wide <- health_df %>%
  # column names = years
  pivot_wider(names_from = Years,
              # rows = Family
              values_from = Health,
              # if there is no Family score, write 'NA'
              values_fill = NA)

# get mean health score of all years
health_wide$mean_score = rowMeans(health_wide[,c(2:6)], na.rm = TRUE)

# round decimals to 2 digits 
health_wide <- health_wide %>%  
  mutate_if(is.numeric,
            round,
            digits = 2)

print(health_wide)
```


# Happiness map
We create a leaflet map were we set the view to 0,0 to get a view of the whole world. 

We then set the labels to the pop-up boxes where we want the HPI score, GDP score, Family score and Health score to be displayed for each country. We uses addPolygons to do this. 

We also use addLegend in order to color the map by the HPI values. 


We get the countries map data set from [this](https://datahub.io/core/geo-countries#r) page. 

```{r world countries, eval=FALSE, echo=TRUE}
countries_world <- readOGR("../input/countries.geojson")
```


First we need to find out if any of the countries in our selfmade data frames are not in the countries_world data frame. 

```{r not in function, eval=FALSE, echo=TRUE}
# function for finding 'not in'
'%nin%' <- Negate('%in%')

# check missing names
HPI_wide$Countries[HPI_wide$Countries  %nin% countries_world$ADMIN]
```


We then need to rename the mismatching countries, so they have the same spelling. 

```{r rename countries, eval=FALSE, echo=TRUE}
# happiness
HPI_wide <- HPI_wide %>%
    mutate(Countries = recode(Countries, 
                              'United States' = 'United States of America',
                              'North Cyprus' = 'Northern Cyprus',
                              'Congo (Kinshasa)' = 'Democratic Republic of the Congo',
                              'Congo (Brazzaville)' = 'Republic of Congo'))

# economy
GDP_wide <- GDP_wide %>%
    mutate(Countries = recode(Countries, 
                              'United States' = 'United States of America',
                              'North Cyprus' = 'Northern Cyprus',
                              'Congo (Kinshasa)' = 'Democratic Republic of the Congo',
                              'Congo (Brazzaville)' = 'Republic of Congo'))

# social support
family_wide <- family_wide %>%
    mutate(Countries = recode(Countries, 
                              'United States' = 'United States of America',
                              'North Cyprus' = 'Northern Cyprus',
                              'Congo (Kinshasa)' = 'Democratic Republic of the Congo',
                              'Congo (Brazzaville)' = 'Republic of Congo'))

# health
health_wide <- health_wide %>%
    mutate(Countries = recode(Countries, 
                              'United States' = 'United States of America',
                              'North Cyprus' = 'Northern Cyprus',
                              'Congo (Kinshasa)' = 'Democratic Republic of the Congo',
                              'Congo (Brazzaville)' = 'Republic of Congo'))
```


Some of the mismatching countries does not appear in the countries_world data set and we will therefore delete the rows in our selfmade data frame.

```{r delete rows, eval=FALSE, echo=TRUE}
HPI_wide <- HPI_wide[-c(146, 91, 162, 165, 166, 167, 169, 108, 87, 72),]
GDP_wide <- GDP_wide[-c(146, 91, 162, 165, 166, 167, 169, 108, 87, 72),]
health_wide <- health_wide[-c(146, 91, 162, 165, 166, 167, 169, 108, 87, 72),]
family_wide <- family_wide[-c(146, 91, 162, 165, 166, 167, 169, 108, 87, 72),]
```


We now want to create a tibble data frame with all the 255 countries from the countries_world data set. 

```{r tibble data frames, eval=FALSE, echo=TRUE}
countries_world_HPI <- tibble(Countries = countries_world$ADMIN)
countries_world_GDP <- tibble(Countries = countries_world$ADMIN)
countries_world_family <- tibble(Countries = countries_world$ADMIN)
countries_world_health <- tibble(Countries = countries_world$ADMIN)
```


Then we want to join the 255 countries into our four selfmade data frames in order to make sure our polygon data and data frames has the same length. 

```{r join countries, eval=FALSE, echo=TRUE}
HPI_polygon_countries <- countries_world_HPI %>% 
  left_join(HPI_wide, by = "Countries", keep = TRUE) %>% 
  distinct(Countries.x,.keep_all = TRUE) 

GDP_polygon_countries <- countries_world_GDP %>% 
  left_join(GDP_wide, by = "Countries", keep = TRUE) %>% 
  distinct(Countries.x,.keep_all = TRUE)

family_polygon_countries <- countries_world_family %>% 
  left_join(family_wide, by = "Countries", keep = TRUE) %>% 
  distinct(Countries.x,.keep_all = TRUE) 

health_polygon_countries <- countries_world_health %>% 
  left_join(health_wide, by = "Countries", keep = TRUE) %>% 
  distinct(Countries.x,.keep_all = TRUE)
```


Define labels for the countries in our map.

```{r set labels, eval=FALSE, echo=TRUE}
# set label names
labels <- sprintf(
  "<strong> %s</strong><br/><strong style=color:rgb(245,189,31);>
  HPI score: </strong> %s <br/><strong style=color:rgb(245,189,31);>
  GDP score: </strong> %s <br/><strong style=color:rgb(245,189,31);>
  Family score: </strong> %s <br/><strong style=color:rgb(245,189,31);>
  Health score: </strong> %s",
  HPI_polygon_countries$Countries.x,
  HPI_polygon_countries$mean_score,
  GDP_polygon_countries$mean_score,
  family_polygon_countries$mean_score,
  health_polygon_countries$mean_score
  ) %>% lapply(htmltools::HTML)
```


Define palette colors for our legend in the map.

```{r palette colors, eval=FALSE, echo=TRUE}
# define palette colors 
pal <- colorNumeric("YlOrRd", 
                    domain = HPI_polygon_countries$mean_score, 
                    na.color = '#388F72')
```


Now we create our beautiful leaflet map.

```{r leaflet map, eval=FALSE, echo=TRUE}
# create leaflet map
map <- leaflet(countries_world) %>%
  addTiles() %>%
  # set view to 0,0 and zoom 1
  setView(lng = 0, lat =  0, zoom = 1) %>% 
  
  # add tiles with mapbox
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = Sys.getenv('MAPBOX_ACCESS_TOKEN'))) %>% 
  
  # add polygons
  addPolygons(
    fillColor = ~pal(HPI_polygon_countries$mean_score),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE),
  
    # add labels 
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto")) %>% 

  # add legend with HPI scores
  addLegend(pal = pal, 
            values = ~HPI_polygon_countries$mean_score, 
            opacity = 0.7, 
            title = "HPI score",
            position = "bottomright")

# save map as a html document 
saveWidget(map, "../output/HPI_map.html", selfcontained = TRUE)
```


# Temperature map
In order to investigate if there is a correlation between happiness and weather, we create a temperature map, so we can make a visual comparison between the map of happiness and the temperature map. 

```{r temperature map, eval=FALSE, echo=TRUE}
# loading raster data "WorldClimate"
climate <- getData(name = "worldclim", var = "bio", res = 10)

# saving the column BIO1, which is the Annual Mean Temperature 
mean_temp <- climate$bio1

# the Worldclim temperature dataset is stated to have a gain of 0.1, which means that it must be multiplied by 0.1 to convert back to degrees Celsius
gain(mean_temp) <- 0.1

# plot map 
plot(mean_temp, main = "Annual Mean Temperature  (in celcius)")
```


# Compare temperature and happiness

First we need to find the top 10 happiest countries.

```{r top 10 HPI, eval=FALSE, echo=TRUE}
# sort by mean score in descending order
top10_hpi <- head(HPI_wide[order(-HPI_wide$mean_score),c(1,7)], n=10)

# sort by alphabetically 
hpi_sort <- top10_hpi[order(top10_hpi$Countries), ]

print(top10_hpi)
```


As we can see the top 10 most happy countries is Denmark, Norway, Finland, Switzerland, Iceland, Netherlands, Canada, Sweden, New Zealand and Australia. 


## Top 10 happiest countries' economy, social support and health
We now want  to take a look at the top 10 countries economy, social support and health to see if they are similar. 


We start with creating a function for finding the top 10 countries.

```{r look for, eval=FALSE, echo=TRUE}
look.for <- c('Denmark', 'Norway', 'Finland', 'Switzerland', 'Iceland', 'Netherlands', 'Canada', 'Sweden', 'New Zealand', 'Australia')
```


We then find the GDP per capita for the top 10 countries and sort them alphabetically.

```{r GDP top10, eval=FALSE, echo=TRUE}
# gdp for top 10
top10_gdp <- GDP_polygon_countries[GDP_polygon_countries$Countries.x %in% look.for, ]

# sort alphabetically
gdp_sort <- top10_gdp[order(top10_gdp$Countries.x), ]
```


... And the social support score.

```{r family top10, eval=FALSE, echo=TRUE}
# family for top 10
top10_family <- family_polygon_countries[family_polygon_countries$Countries.x %in% look.for, ]

# sort alphabetically
family_sort <- top10_family[order(top10_family$Countries.x), ]
```


... And the health score.

```{r health top10, eval=FALSE, echo=TRUE}
# health for top 10
top10_health <- health_polygon_countries[health_polygon_countries$Countries.x %in% look.for, ]

# sort alphabetically
health_sort <- top10_health[order(top10_health$Countries.x), ]
```


We then create a data frame with four values for each of the top 10 countries. 
```{r top10 new data frame, eval=FALSE, echo=TRUE}
# bind columns 
top10_df <- cbind(hpi_sort$Countries, hpi_sort$mean_score, gdp_sort$mean_score, family_sort$mean_score, health_sort$mean_score)

# convert to data frame
top10_df <- as.data.frame(top10_df)

# rename columns
top10_df <- top10_df %>% 
  rename(
    Countries = V1,
    'HPI score' = V2,
    'GDP score' = V3, 
    'Family score' = V4,
    'Health score' = V5
    )

print(top10_df)
```



## Temperature for the top 10 happiest countries
Find temperature for the 10 happiest countries to see if the climate is about the same for the 10 countries.

We have used the longitude and latitude for the capitals in the top 10 countries. We have found the coordinates on [this](https://www.findlatitudeandlongitude.com/) page. 

```{r temperature for top 10 HPI, eval=FALSE, echo=TRUE}
# get first column in climate dataset with annual temperature 
climate_temp <- climate[[c(1)]]
names(climate_temp) <- c("Temp")

# convert to celsisius
gain(climate_temp) <- 0.1

# get latitude for top 10
lat <- c(55.676098, 
         59.913868,
         60.186826,
         46.947974,
         64.146582,
         52.367984,
         45.42153,
         59.329323,
         -41.28646,
         -35.280937)

# get longitude for top 10
long <- c(12.568337,
          10.752245,
          24.935301,
          7.447447,
          -21.942635,
          4.903561,
          -75.697193,
          18.06858,
          174.776236,
          149.130009) 

# coordinates
coords <- data.frame(x=long,y=lat)
# spatial points
points <- SpatialPoints(coords, proj4string = climate_temp@crs)

# extract values
values <- extract(climate_temp,points)

# combine columns to new data frame 
top10_temp <- cbind.data.frame(coordinates(points),values)

# rename columns
top10_temp <- top10_temp %>% 
  rename(
    Longitude = x,
    Latitude = y, 
    'Annual mean temperature' = values
    )

# create new column with country names
top10_temp$Country <- c('Denmark', 'Norway', 'Finland', 'Switzerland', 'Iceland', 'Netherlands', 'Canada', 'Sweden', 'New Zealand', 'Australia')

# change column order
top10_temp <- top10_temp %>% relocate(Country, .before = Longitude)

print(top10_temp)
```


Plot the top 10 countries on temperature map 

```{r plot, eval=FALSE, echo=TRUE}
plot(climate_temp[[1]])
plot(points,add=T)
```

